---
title: "Report: RNA-Seq counts summary"
author: "`r sessionInfo()$platform`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
   rmd: "report.Rmd"
output:
  html_document:
    theme: cosmo
    toc: yes
    toc_depth: 2
    number_sections: yes
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Background

This report summarizes the results from the read counting step.
It provides quality control plots and statistics for each condition and replicate.
It is recommended to check this report when problems or questions about the DE analysis appear.

# Requirements

- loading libraries `tidyverse`, `grDevices`

```{r, warning = FALSE}
suppressPackageStartupMessages({
    library(grDevices)
    library(tidyverse)
})
```

```{r, echo = FALSE}
# messages
messages <- c()

figure_dir <- gsub("report.html$", "figures/", snakemake@output$html)
if (!dir.exists(figure_dir)) {
  dir.create(figure_dir)
  messages <- append(messages, paste0("created figure dir: ", figure_dir))
}

# custom ggplot2 theme that is reused for all later plots
custom_colors <- c(
  "#E7298A", "#66A61E", "#E6AB02", "#7570B3",
  "#B3B3B3", "#1B9E77", "#D95F02", "#A6761D"
)

custom_range <- function(n = 5, quantitative = FALSE) {
  if (quantitative) {
    colorRampPalette(custom_colors[c(5, 3, 4)])(n)
  } else {
    colorRampPalette(custom_colors)(n)
  }
}

custom_theme <- function(
    base_size = 12, base_line_size = 1.0,
    base_rect_size = 1.0, ...) {
  theme_light(
    base_size = base_size,
    base_line_size = base_line_size,
    base_rect_size = base_rect_size
  ) +
    theme(
      plot.margin = margin(10, 10, 10, 10),
      title = element_text(colour = grey(0.4), size = 10),
      plot.subtitle = element_text(colour = grey(0.4), size = 9),
      axis.ticks.length = unit(0.2, "cm"),
      axis.ticks = element_line(
        colour = grey(0.4),
        linetype = "solid", lineend = "round"
      ),
      axis.text.x = element_text(colour = grey(0.4), size = 10),
      axis.text.y = element_text(colour = grey(0.4), size = 10),
      panel.grid.major = element_line(
        linewidth = 0.6,
        linetype = "solid", colour = grey(0.9)
      ),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(
        linetype = "solid",
        colour = grey(0.4), fill = NA, linewidth = 1.0
      ),
      panel.background = element_blank(),
      strip.background = element_blank(),
      strip.text = element_text(
        colour = grey(0.4), size =
          10, margin = unit(rep(3, 4), "points")
      ),
      legend.text = element_text(colour = grey(0.4), size = 10),
      legend.title = element_blank(),
      legend.background = element_blank(),
      ...
    )
}

# set graphical parameter for subfigure labels
list_fontpars <- list(face = "plain", size = 14)

# function to export an image as svg and png
save_plot <- function(
    pl, path = figure_dir,
    width = figwidth, height = figheight, index = NA) {
  pl_name <- deparse(substitute(pl))
  if (!is.na(index)) pl_name <- paste0(pl_name, "_", index)
  svg(
    filename = paste0(path, pl_name, ".svg"),
    width = width, height = height
  )
  print(pl)
  dev.off()
  png(
    filename = paste0(path, pl_name, ".png"),
    width = width * 125, height = height * 125, res = 120
  )
  print(pl)
  invisible(capture.output(dev.off()))
}
```

# Sample overview

## Data import

- imported samplesheet `r snakemake@input[["samplesheet"]]`
- imported raw counts file `r snakemake@input[["merged_counts"]]`
- imported filtered counts file `r snakemake@input[["filtered_counts"]]`

```{r}
# table import
samplesheet <- snakemake@input[["samplesheet"]]
raw_counts <- snakemake@input[["merged_counts"]]
filt_counts <- snakemake@input[["filtered_counts"]]

df_samples <- read_tsv(samplesheet, show_col_types = FALSE)
df_raw_counts <- read_csv(raw_counts, show_col_types = FALSE) %>%
  pivot_longer(cols = -c(Geneid), names_to = "sample", values_to = "n_reads")
df_filt_counts <- read_csv(filt_counts, show_col_types = FALSE) %>%
  pivot_longer(cols = -c(Geneid), names_to = "sample", values_to = "n_reads")


# list of samples + generic options
figwidth <- 9
figheight <- round(1 + (length(nrow(df_samples)) / 4))
figheight2 <- 1.6 * figheight
```

## Samples

- overview of samples
- read metric are aggregated per gene/feature

```{r}
df_summary <- bind_rows(
  list(raw = df_raw_counts, filtered = df_filt_counts),
  .id = "type"
) %>%
  dplyr::group_by(type, sample) %>%
  dplyr::summarize(
    total_reads = sum(n_reads, na.rm = TRUE),
    min_reads = min(n_reads, na.rm = TRUE),
    mean_reads = mean(n_reads, na.rm = TRUE),
    median_reads = median(n_reads, na.rm = TRUE),
    max_reads = max(n_reads, na.rm = TRUE),
  ) %>%
  arrange(desc(type)) %>%
  mutate(type = factor(type, c("filtered", "raw")))

print(df_summary)
```

## QC metrics

### Mapped reads per sample

- figure shows the total number of mapped reads
- reads are shown before and after filtering

```{r, fig.width = figwidth, fig.height = figheight, warning = FALSE}
plot_total_mapped_reads <- df_summary %>%
    ggplot(aes(x = sample, y = total_reads, fill = type)) +
    coord_flip() +
    geom_col(alpha = 0.8, position = "identity") +
    labs(x = "", y = "total number of mapped reads") +
    custom_theme() +
    scale_fill_manual(values = custom_colors[c(1, 5)])

save_plot(plot_total_mapped_reads, width = figwidth, height = figheight)
print(plot_total_mapped_reads)
```

### Mean reads per gene

- figure shows the mean number of reads mapped per gene and sample

```{r, fig.width = figwidth, fig.height = figheight, warning = FALSE}
plot_mean_mapped_reads <- df_summary %>%
    ggplot(aes(x = sample, y = mean_reads, fill = type)) +
    coord_flip() +
    geom_col(alpha = 0.8, position = "identity") +
    labs(x = "", y = "mean number of mapped reads") +
    custom_theme() +
    scale_fill_manual(values = custom_colors[c(1, 5)])

save_plot(plot_mean_mapped_reads, width = figwidth, height = figheight)
print(plot_mean_mapped_reads)
```

### Max and min reads per gene

- figure shows the maximum and minimum number of reads mapped per gene and sample
- only filtered counts are shown

```{r, fig.width = figwidth, fig.height = figheight, warning = FALSE}
plot_min_max_reads <- df_summary %>%
  filter(type == "filtered") %>%
    ggplot(aes(x = sample)) +
    geom_point(aes(y = min_reads), shape = 19, size = 2, color = custom_colors[1]) +
    geom_point(aes(y = max_reads), shape = 19, size = 2, color = custom_colors[1]) +
    geom_segment(aes(y = min_reads, yend = max_reads, xend = sample), color = custom_colors[1], linewidth = 1.0) +
    coord_flip() +
    labs(x = "", y = "mean number of mapped reads") +
    custom_theme()

save_plot(plot_min_max_reads, width = figwidth, height = figheight)
print(plot_min_max_reads)
```

### Number of missing gene quantifications per sample

- figure shows the number of missing genes per sample (zero reads)
- only filtered counts are shown

```{r, fig.width = figwidth, fig.height = figheight, warning = FALSE}
plot_missing_quants <- df_filt_counts %>%
  dplyr::group_by(sample) %>%
  dplyr::summarize(`missing gene quantifications` = sum(n_reads == 0)) %>%
  ggplot(aes(x = sample, y = `missing gene quantifications`)) +
  geom_col(fill = custom_colors[1], alpha = 0.8) +
  labs(x = "") +
  coord_flip() +
  custom_theme()

save_plot(plot_missing_quants, width = figwidth, height = figheight)
print(plot_missing_quants)
```

### Read count distribution, violin plot

- figure shows the read count distribution per sample
- genes without mapped reads for the respective sample are removed
- read count is log 10 transformed (0 -> 1, 1 -> 10, 2 -> 100, ...)

```{r, fig.width = figwidth, fig.height = figheight, warning = FALSE}
plot_reads_per_gene_violin <- df_filt_counts %>%
    ggplot(aes(x = log10(n_reads), y = sample)) +
    geom_violin(fill = custom_colors[1], alpha = 0.8, color = NA) +
    labs(y = "", x = expression("log"[10] * " reads per gene")) +
    custom_theme()

save_plot(plot_reads_per_gene_violin, width = figwidth, height = figheight)
print(plot_reads_per_gene_violin)
```

### Read count distribution, histogram

- figure shows the same data as above, only as histogram
- genes without mapped reads for the respective sample are removed
- read count is log 10 transformed (0 -> 1, 1 -> 10, 2 -> 100, ...)

```{r, fig.width = figwidth, fig.height = figheight2, warning = FALSE}
plot_reads_per_gene_hist <- df_filt_counts %>%
    ggplot(aes(x = log10(n_reads))) +
    geom_histogram(fill = custom_colors[1], alpha = 0.7, bins = 30) +
    labs(y = "", x = expression("log"[10] * " reads per gene")) +
    facet_wrap( ~ sample, ncol = 1) +
    custom_theme()

save_plot(plot_reads_per_gene_hist, width = figwidth, height = figheight2)
print(plot_reads_per_gene_hist)
```

### Top 10 most abundant genes by read number

- figure shows top 10 genes ranked by read count
- only filtered counts are shown

```{r, fig.width = figwidth, fig.height = figheight2, warning = FALSE}
plot_top10_genes <- df_filt_counts %>%
    dplyr::group_by(sample) %>%
    dplyr::arrange(sample, desc(n_reads)) %>%
    dplyr::mutate(rank = seq_along(Geneid)) %>%
    dplyr::filter(between(rank, 1, 10)) %>%
    ggplot(aes(x = factor(rank), y = n_reads)) +
    geom_col(fill = custom_colors[1], alpha = 0.7, width = 1) +
    labs(y = "n reads", x = "genes ranked by abundance") +
    facet_wrap(~sample, ncol = 4) +
    custom_theme()

save_plot(plot_top10_genes, width = figwidth, height = figheight2)
print(plot_top10_genes)
```

### Cumulative read count distribution

- figure shows the gene diversity by plotting fraction of reads (%) vs fraction of genes (%)
- the ideal experiment has high diversity and equal distribution of genes for similar samples
- such a distribution would follow the diagonal dashed grey line
- if read distribution (red line) is highly skewed, `% of reads` (y-axis) shows a steep ascent
- this means very few genes contribute to almost all reads

```{r, fig.width = figwidth, fig.height = figheight2, warning = FALSE}
df_auc <- df_filt_counts %>%
    dplyr::arrange(sample, desc(n_reads)) %>%
    dplyr::group_by(sample) %>%
    dplyr::mutate(
        pc_genes = 100 * seq_along(n_reads) / n(),
        pc_reads = 100 * n_reads / (sum(n_reads))
    ) %>%
    dplyr::mutate(pc_genes = round(pc_genes)) %>%
    dplyr::group_by(sample, pc_genes) %>%
    dplyr::summarize(pc_reads = sum(pc_reads), .groups = "drop_last") %>%
    dplyr::mutate(pc_reads = cumsum(pc_reads))

plot_cumulative_read_count <- df_auc %>%
    ggplot(aes(x = pc_genes, y = pc_reads, group = sample)) +
    geom_step(linewidth = 1.0, color = custom_colors[1]) +
    geom_abline(
        slope = 1, intercept = 0, linewidth = 1.0,
        linetype = 2, color = grey(0.5)
    ) +
    lims(x = c(0, 100), y = c(0, 100)) +
    labs(x = "% of genes", y = "% of reads") +
    facet_wrap(~sample, ncol = 4) +
    custom_theme()

save_plot(plot_cumulative_read_count, width = figwidth, height = figheight2)
print(plot_cumulative_read_count)
```

- this table shows the area under curve (AUC) for the line plot above
- an AUC of 0.5 is ideal, an AUC approaching 1.0 is not optimal
- the 'Gini index' is a score between 0 and 1 measuring population equality
- it is defined as the `(AUC - AUC_optimal) / AUC_optimal`
- a Gini score of 0 describes a perfectly equal, a Gini score of 1.0 a perfectly unequal distribution

```{r, warning = FALSE}
calc_auc <- function(x, y) {
    sum(diff(x) * (head(y, -1) + tail(y, -1))) / 2
}
df_auc %>%
    dplyr::summarize(
        auc = calc_auc(pc_genes, pc_reads)/100^2,
        auc_optimal = 0.5,
        gini = (auc - 0.5) / 0.5
    ) %>%
    mutate(quality = case_when(
        gini <= 0.33 ~ "low inequality (G < 0.33)",
        gini > 0.33 & gini <= 0.66 ~ "intermediate inequality (0.33 < G < 0.66)",
        gini > 0.66 ~ "high inequality (G > 0.66)"
    ))
```


### Sample and replicate correlation coefficent (R)

- figure shows heat map with correlation coefficient R (-1 < R < 1)
- correlation coefficient shows how strongly read abundance is correlated

```{r, fig.width = 7.5, fig.height = 7, warning = FALSE}
df_correlation <- df_filt_counts %>%
    tidyr::pivot_wider(names_from = "sample", values_from = "n_reads") %>%
    dplyr::select(-Geneid) %>%
    cor()

plot_replicate_correlation <- df_correlation %>%
    dplyr::as_tibble() %>%
    dplyr::mutate(sample1 = colnames(.)) %>%
    tidyr::pivot_longer(
        cols = !sample1,
        names_to = "sample2", values_to = "cor_coef"
    ) %>%
    ggplot(aes(x = sample1, y = sample2, fill = cor_coef)) +
    geom_tile() +
    geom_text(color = grey(0.4), aes(label = round(cor_coef, 2))) +
    custom_theme() +
    labs(title = "", x = "", y = "") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    scale_fill_gradientn(
        colours = c(custom_colors[1], grey(0.9), custom_colors[2]),
        limits = c(-1, 1)
    )

write.csv(df_correlation, file = "correlation_table.csv")
save_plot(plot_replicate_correlation, width = 7.5, height = 7)
print(plot_replicate_correlation)
```

### Sample and replicate similarity with PCA

- this figure shows sample similarity similar to above figure
- uses principal component analysis (PCA) to reduce the multidimensional data to 3 main dimensions
- plotted are principal component 1 (x-axis), 2 (y-axis) and 3 (point size)
- replicates for similar conditions should cluster together
- outliers should be easily visible

```{r, fig.width = 7, fig.height = 7, warning = FALSE}
pca_result <- df_filt_counts %>%
    tidyr::pivot_wider(names_from = "sample", values_from = "n_reads") %>%
    dplyr::select(-Geneid) %>%
    as.matrix() %>%
    t() %>%
    replace(., is.na(.), 0) %>%
    prcomp()

df_PCA <- pca_result$x %>%
    as_tibble(rownames = "sample") %>%
    left_join(
      dplyr::select(df_samples, condition, replicate, sample),
      by = join_by(sample)
    )

plot_replicate_pca <- df_PCA %>%
    ggplot(aes(x = PC1, y = PC2, size = PC3, color = condition, label = replicate)) +
    geom_point(alpha = 0.8) +
    geom_text(
      size = 4, show.legend = FALSE, 
      nudge_y = diff(range(df_PCA$PC2))/30,
      nudge_x = diff(range(df_PCA$PC1))/30) +
    labs(
        title = "PCA, first three principal components",
        subtitle = "Point size: PC3, color: condition, number: replicate.", x = "PC1", y = "PC2"
    ) +
    custom_theme(legend.position = "bottom", aspect.ratio = 1) +
    scale_color_manual(values = colorRampPalette(custom_colors)(length(unique(df_PCA$condition))))

save_plot(plot_replicate_pca, width = 7, height = 7)
print(plot_replicate_pca)
```

# About this report

## Workflow

This report was automatically generated by the [snakemake-bacterial-rnaseq-deseq](https://github.com/MPUSP/snakemake-bacterial-rnaseq-deseq) workflow.

For issues, bugs, and feature requests please use the workflow's [github page](https://github.com/MPUSP/snakemake-bacterial-rnaseq-deseq/issues).

For all other feedback, contact the author(s):

- Michael Jahn, PhD: jahn@mpusp.mpg.de
- Rina Ahmed-Begrich, PhD: begrich@mpusp.mpg.de

The workflow is developed at the [Max Planck Unit for the Science of Pathogens](https://www.mpusp.mpg.de/), Berlin, Germany.

## Configuration

The pipeline configuration file is available at: `config/config.yml`.

```{r, echo = FALSE}
list_config <- read_lines("config/config.yml")
list_config <- gsub("\"", "", list_config)
print(list_config)
```

## Data accessability

Input tables are listed at the top of the report.

## Session Info

Link to the R markdown source that was used to generate this report:

<a download="report.Rmd" href="`r base64enc::dataURI(file = params$rmd, mime = 'text/rmd', encoding = 'base64')`">R Markdown source file</a>

Session info (R base and package versions):

```{r, echo = FALSE}
sessionInfo()
```

```{r, echo = FALSE}
messages <- append(messages, "finished writing HTML report successfully")
write_lines(
  paste0("HTML_REPORT: ", messages),
  file = "results/deseq2/counts_report.log"
)
```
