---
title: "Report: RNA-Seq DESeq2 summary"
author: "`r sessionInfo()$platform`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  rmd: "report.Rmd"
output:
  html_document:
    theme: cosmo
    toc: yes
    toc_depth: 2
    number_sections: yes
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# Background

This report summarizes the results from running [DESeq2](https://bioconductor.org/packages/release/bioc/html/DESeq2.html),
a package for the analysis of *differentially expressed* (DE) genes.
The full analysis of RNA-Seq data was performed using two Snakemake workflows:

1. [Snakemake workflow for bacterial RNA-Seq data processing](https://github.com/MPUSP/snakemake-bacterial-rnaseq-processing)
2. [Snakemake workflow for bacterial RNA-Seq data analysis with DESeq2](https://github.com/MPUSP/snakemake-bacterial-rnaseq-deseq)

The first workflow processes raw RNA-Seq reads and generates a table of read counts per gene.
The second workflow uses this table to perform the DE analysis with DESeq2.
Results from the first workflow are not included in this report, but can be found in `results/deseq2/counts_report.html`.

# Requirements

- loading libraries `tidyverse`, `grDevices`, `DESeq2`, `ggrepel`

```{r}
suppressPackageStartupMessages({
  library(grDevices)
  library(tidyverse)
  library(DESeq2)
  library(ggrepel)
})
```

```{r}
# messages
messages <- c()

figure_dir <- gsub("report.html$", "figures/", snakemake@output$html)
if (!dir.exists(figure_dir)) {
  dir.create(figure_dir)
  messages <- append(messages, paste0("created figure dir: ", figure_dir))
}

# custom ggplot2 theme that is reused for all later plots
custom_colors <- c(
  "#E7298A", "#66A61E", "#E6AB02", "#7570B3",
  "#B3B3B3", "#1B9E77", "#D95F02", "#A6761D"
)

custom_range <- function(n = 5, quantitative = FALSE) {
  if (quantitative) {
    colorRampPalette(custom_colors[c(5, 3, 4)])(n)
  } else {
    colorRampPalette(custom_colors)(n)
  }
}

custom_theme <- function(
    base_size = 12, base_line_size = 1.0,
    base_rect_size = 1.0, ...) {
  theme_light(
    base_size = base_size,
    base_line_size = base_line_size,
    base_rect_size = base_rect_size
  ) +
    theme(
      plot.margin = margin(10, 10, 10, 10),
      title = element_text(colour = grey(0.4), size = 10),
      plot.subtitle = element_text(colour = grey(0.4), size = 9),
      axis.ticks.length = unit(0.2, "cm"),
      axis.ticks = element_line(
        colour = grey(0.4),
        linetype = "solid", lineend = "round"
      ),
      axis.text.x = element_text(colour = grey(0.4), size = 10),
      axis.text.y = element_text(colour = grey(0.4), size = 10),
      panel.grid.major = element_line(
        linewidth = 0.6,
        linetype = "solid", colour = grey(0.9)
      ),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(
        linetype = "solid",
        colour = grey(0.4), fill = NA, linewidth = 1.0
      ),
      panel.background = element_blank(),
      strip.background = element_blank(),
      strip.text = element_text(
        colour = grey(0.4), size = 10, 
        margin = margin(3, 3, 3, 3)
      ),
      legend.text = element_text(colour = grey(0.4), size = 10),
      legend.title = element_blank(),
      legend.background = element_blank(),
      ...
    )
}

# set graphical parameter for subfigure labels
list_fontpars <- list(face = "plain", size = 14)

# function to export an image as svg and png
save_plot <- function(
    pl, path = figure_dir,
    width = figwidth, height = figheight, index = NA) {
  pl_name <- deparse(substitute(pl))
  if (!is.na(index)) pl_name <- paste0(pl_name, "_", index)
  svg(
    filename = paste0(path, pl_name, ".svg"),
    width = width, height = height
  )
  print(pl)
  dev.off()
  png(
    filename = paste0(path, pl_name, ".png"),
    width = width * 125, height = height * 125, res = 120
  )
  print(pl)
  invisible(capture.output(dev.off()))
}
```

# Sample overview

## Data import

- imported samplesheet `r snakemake@input[["samplesheet"]]`
- imported DESeq2 results raw data `r snakemake@input[["deseq_data"]]`
- imported DESeq2 results table `r snakemake@input[["deseq_results"]]`


```{r}
df_samples <- read_tsv(snakemake@input[["samplesheet"]], show_col_types = FALSE)
df_deseq <- read_csv(snakemake@input[["deseq_results"]], show_col_types = FALSE) %>%
  mutate(significance = factor(significance, levels = c("significant", "not significant")))
load(snakemake@input[["deseq_data"]])

threshold_pval <- snakemake@config$deseq2$threshold_pval
if (is.null(threshold_pval) || is.na(threshold_pval)) threshold_pval <- 0.05
threshold_log2fc <- snakemake@config$deseq2$threshold_log2fc
if (is.null(threshold_log2fc) || is.na(threshold_log2fc)) threshold_log2fc <- 1.0

figwidth <- 9
figheight <- ceiling(1.0 + nrow(df_samples) / 4)
figheight2 <- 1.6 * figheight

df_samples
```

## Differential gene expression

- DESeq2 allows an unlimited number of comparisons between arbitrary conditions
- for this to work, every condition must include more than 1 replicate
- the following sections show results for all valid comparisons

### User-defined comparisons

- comparisons are defined in the sample sheet using the `comparison` column
- each condition will be contrasted against all listed reference conditions in this column
- note: redundant comparisons like A vs B and B vs A are not filtered, it can be meaningful to look at differences from both sides

```{r}
df_comparisons <- df_samples %>%
  group_by(condition, comparison) %>%
  dplyr::summarize(`number of replicates` = n(), .groups = "drop")

df_comparisons
```

### PCA (using `baseMean` from DESeq2)

- this figure shows sample similarity between conditions
- replicate information is not part of this analysis as it is already aggregated by `DESeq2`
- uses principal component analysis (PCA) to reduce the multidimensional data to 3 main dimensions
- plotted are principal component 1 (x-axis), 2 (y-axis) and 3 (point size)
- similar conditions should cluster together
- outliers should be easily visible

```{r, fig.width = 7, fig.height = 7}
df_pca <- plotPCA(vst(deseq_data, nsub = 100), intgroup = "condition", returnData = TRUE)

plot_pca <- df_pca %>%
  ggplot(aes(x = PC1, y = PC2, color = condition, label = name)) +
  geom_point(alpha = 0.8) +
  geom_text(
    size = 4, show.legend = FALSE,
    nudge_y = diff(range(df_pca$PC2)) / 30,
    nudge_x = diff(range(df_pca$PC1)) / 30
  ) +
  labs(
    title = "PCA, first two principal components",
    subtitle = "Point size: PC3, color: condition, number: replicate.", x = "PC1", y = "PC2"
  ) +
  custom_theme(legend.position = "bottom", aspect.ratio = 1) +
  scale_color_manual(values = colorRampPalette(custom_colors)(length(unique(df_pca$condition))))

save_plot(plot_pca, width = 7, height = 7)
print(plot_pca)
```

### One-by-one comparison, Volcano plot {.tabset}

- shows all genes as dots
- x-axis: log2 fold change (DESeq2 column `log2FoldChange`)
- y-axis: -log10 adjusted p-value (DESeq2 column `padj`)

---

**Overview figure**

```{r, fig.width = 8, fig.height = 8}
plot_volcano_overview <- ggplot(
  data = df_deseq,
  aes(x = log2FoldChange, y = -log10(padj))
) +
  geom_point(aes(color = significance), alpha = 0.6, size = 1) +
  labs(
    title = "Volcano Plot",
    subtitle = paste0("significance: p ≤ ", threshold_pval, ", log2FC ≥ ", threshold_log2fc),
    x = "Log2 Fold Change",
    y = "-log10 p-value"
  ) +
  lims(
    x = {if (any(abs(range(df_deseq$log2FoldChange)) > 2)) {range(df_deseq$log2FoldChange)} else {c(-2, 2)}},
    y = {if (max(-log10(df_deseq$padj)) > 3) {c(0, max(-log10(df_deseq$padj)))} else {c(0, 3)}}
  ) +
  theme(
    legend.title = element_blank(),
    legend.position = "bottom",
    aspect.ratio = 1
  ) +
  facet_wrap(~ paste0(condition, " ~ ", reference), ncol = 4) +
  custom_theme(legend.position = "bottom", aspect.ratio = 1) +
  scale_color_manual(values = custom_colors[c(1, 5)])

save_plot(plot_volcano_overview, width = 8, height = 8)
print(plot_volcano_overview)
```

---

**One-by-one comparison**

```{r, results = 'asis', fig.width = 8, fig.height = 8}
comparisons <- df_comparisons %>%
  group_by(condition, comparison) %>%
  group_split()

for (comp in comparisons) {
  cat("\n\n#### ", comp$condition, " - ", comp$comparison, " {-}\n\n")
  plot_volcano <- df_deseq %>%
    filter(condition == comp$condition, reference == comp$comparison) %>%
    mutate(trivial_name = ifelse(significance == "significant", trivial_name, "")) %>%
    ggplot(aes(x = log2FoldChange, y = -log10(padj), color = significance, label = trivial_name)) +
    geom_point(alpha = 0.6) +
    geom_label_repel(size = 3, box.padding = 0.2, max.overlaps = 10) +
    labs(
      title = paste0("Volcano Plot: ", comp$condition, " - ", comp$comparison),
      subtitle = paste0("significance: p ≤ ", threshold_pval, ", log2FC ≥ ", threshold_log2fc),
      x = "Log2 Fold Change",
      y = "-log10 p-value"
    ) +
    lims(
      x = {if (any(abs(range(df_deseq$log2FoldChange)) > 2)) {range(df_deseq$log2FoldChange)} else {c(-2, 2)}},
      y = {if (max(-log10(df_deseq$padj)) > 3) {c(0, max(-log10(df_deseq$padj)))} else {c(0, 3)}}
    ) +
    theme(
      legend.title = element_blank(),
      legend.position = "bottom",
      aspect.ratio = 1
    ) +
    custom_theme(legend.position = "bottom", aspect.ratio = 1) +
    scale_color_manual(values = custom_colors[c(1, 5)])

  print(plot_volcano)
  save_plot(
    plot_volcano, width = 8, height = 8,
    index = paste0(comp$condition, "_vs_", comp$comparison)
  )
}
```

### One-by-one comparison, MA pLot  {.tabset}

- shows all genes as dots
- x-axis: mean expression across all samples (DESeq2 column `baseMean`)
- y-axis: log2 fold change (DESeq2 column `log2FoldChange`)
- how the MA plot is interpreted: 
  - points high y-values are probably differentially expressed
  - points with low x-values have low expression and are less reliable

**Overview figure**

```{r, fig.width = 8, fig.height = 8}
plot_ma_overview <- ggplot(
  data = df_deseq,
  aes(x = baseMean, y = log2FoldChange)
) +
  geom_point(aes(color = significance), alpha = 0.6, size = 1) +
  labs(
    title = "MA Plot",
    subtitle = paste0("significance: p ≤ ", threshold_pval, ", log2FC ≥ ", threshold_log2fc),
    x = "Mean Expression",
    y = "Log2 Fold Change"
  ) +
  ylim({if (any(abs(range(df_deseq$log2FoldChange)) > 2)) {range(df_deseq$log2FoldChange)} else {c(-2, 2)}}) +
  scale_x_continuous(
    limits = {if (max(df_deseq$baseMean) > 10) {c(0.1, max(df_deseq$baseMean))} else {c(0.1, 10)}},
    transform = "log10"
  ) +
  theme(
    legend.title = element_blank(),
    legend.position = "bottom",
    aspect.ratio = 1
  ) +
  facet_wrap(~ paste0(condition, " ~ ", reference), ncol = 4) +
  custom_theme(legend.position = "bottom", aspect.ratio = 1) +
  scale_color_manual(values = custom_colors[c(1, 5)])

save_plot(plot_ma_overview, width = 8, height = 8)
print(plot_ma_overview)
```

**One-by-one comparison**

```{r, results = 'asis', fig.width = 8, fig.height = 8}
for (comp in comparisons) {
  cat("\n\n#### ", comp$condition, " - ", comp$comparison, " {-}\n\n")
  plot_ma <- df_deseq %>%
    filter(condition == comp$condition, reference == comp$comparison) %>%
    mutate(trivial_name = ifelse(significance == "significant", trivial_name, "")) %>%
    ggplot(aes(x = baseMean, y = log2FoldChange, color = significance, label = trivial_name)) +
    geom_point(alpha = 0.6, size = 1) +
    geom_label_repel(size = 3, box.padding = 0.2, max.overlaps = 10) +
    labs(
      title = paste0("Volcano Plot: ", comp$condition, " - ", comp$comparison),
      subtitle = paste0("significance: p ≤ ", threshold_pval, ", log2FC ≥ ", threshold_log2fc),
      x = "Mean Expression",
      y = "Log2 Fold Change"
    ) +
    ylim({if (any(abs(range(df_deseq$log2FoldChange)) > 2)) {range(df_deseq$log2FoldChange)} else {c(-2, 2)}}) +
    scale_x_continuous(
      limits = {if (max(df_deseq$baseMean) > 10) {c(0.1, max(df_deseq$baseMean))} else {c(0.1, 10)}},
      transform = "log10"
    ) +
    theme(
      legend.title = element_blank(),
      legend.position = "bottom",
      aspect.ratio = 1
    ) +
    custom_theme(legend.position = "bottom", aspect.ratio = 1) +
    scale_color_manual(values = custom_colors[c(1, 5)])

  print(plot_ma)
  save_plot(
    plot_ma, width = 8, height = 8,
    index = paste0(comp$condition, "_vs_", comp$comparison)
  )
}
```

# About this report

## Workflow

This report was automatically generated by the [snakemake-bacterial-rnaseq-deseq](https://github.com/MPUSP/snakemake-bacterial-rnaseq-deseq) workflow.

For issues, bugs, and feature requests please use the workflows [github page](https://github.com/MPUSP/snakemake-bacterial-rnaseq-deseq/issues).

For all other feedback, contact the author(s):

- Michael Jahn, PhD: jahn@mpusp.mpg.de
- Rina Ahmed-Begrich, PhD: begrich@mpusp.mpg.de

The workflow is developed at the [Max Planck Unit for the Science of Pathogens](https://www.mpusp.mpg.de/), Berlin, Germany.

## Configuration

The pipeline configuration file is available at: `config/config.yml`.

```{r}
list_config <- read_lines("config/config.yml")
list_config <- gsub("\"", "", list_config)
print(list_config)
```

## Data accessability

Input tables are listed at the top of the report.

## Session Info

Link to the R markdown source that was used to generate this report:

<a download="report.Rmd" href="`r base64enc::dataURI(file = params$rmd, mime = 'text/rmd', encoding = 'base64')`">R Markdown source file</a>

Session info (R base and package versions):

```{r}
sessionInfo()
```

```{r}
messages <- append(messages, "finished writing HTML report successfully")
write_lines(
  paste0("HTML_REPORT: ", messages),
  file = "results/deseq2/deseq2_report.log"
)
```
